Supplementary Information

### 2021-05-07 doc/manuscript/ForestGEO_treerings.Rmd
finish Sean feedback, misc

Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021], and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @rollinson_climate_2021], which may often be best described by nonlinear functions [REFS, @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., @evans_fusing_2017; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and as a result we lack synthetic, global-scale understanding of some of the key questions that may be addressed using this such an approach (Table `r hypotheses`).

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
Over the vast majority of Earth's forested regions, water is the primary climatic factor limiting tree growth, whereas low temperatures are the strongest limitation in some high latitude or high elevation sites [@babst_twentieth_2019]. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the main growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., **REFS**, @buntgen_2500_2011], more in-depth characterization of climate sensitivity across a representative sample of all trees in a forest stand is desirable for efforts to understand forest productivity and its climate responses [@babst_when_2018].
First, the additive or interactive effects of moisture and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020].
We hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`) such that their effects cannot be fully captured by a single metric such as the Standardised Precipitation-Evapotranspiration Index [SPEI, @vicente-serrano_multiscalar_2010].
In addition, based on the predominance of nonlinear climate responses of tree growth over both smaller and larger spatiotemporal scales [e.g., @kumarathunge_acclimation_2019; @banburymorgan_global_2021], we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @trouillier_size_2019; @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].

@@

Here, we develop, and apply to ten forested sites spanning 52 degrees latitude, a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
This approach allows us to address the following questions and associated hypotheses (Table `r hypotheses`) for a globally distributed tree-ring dataset:
(1) How does interannual climate variation shape annual tree growth (ring width, RW)?
(2) How does growth rate vary with DBH?
(3) How have growth rates changed through time?

@@

Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability (PPT or PDF) -- at least up until the long-term mean -- 
and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
These findings are generally consistent with current understanding of global-scale patterns in climate sensitivity [@rozendaal_dendroecology_2011; @babst_twentieth_2019]: the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone). 
Within warmer regions, forests in humid climates can sometimes benefit from warm winter or spring temperatures [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig. `r clim_sens`).
However, the predominantly negative temperature responses (Fig. `r clim_sens`) imply that warmer temperatures are likely to reduce growth across the wide range of forest types and climates represented here.
This growth decrease at high temperatures is probably driven primarily by increased atmospheric dryness (PET or vapor pressure deficit) and ensuant exacerbation of observed water limitations (**REFS**) [@humphrey_soil_2021; @lopez_systemic_2021].
This effect occurs in addition to the effects of precipitation (Fig. `r clim_sens`), highlighting the fact that precipitation and temperature jointly influence drought (**REFS?**), but, critically, tends to occur over shorter time-frames (usually $\le$ 3 mo) than the effects of precipitation (usually $\ge$ 3 mo.; Table `r hypotheses`, Fig. `r clim_sens`).
This suggests that relatively short periods of anomalously high temperatures and atmospheric dryness, themselves caused in large part by soil dryness [@humphrey_soil_2021], add to effects of prolonged periods of reduced precipitation to shape forest drought responses.

Our analysis differed fundamentally from most conventional approaches in testing for non-linear responses of growth to climate, finding that the majority of climate responses were nonlinear (Table `r hypotheses`, Fig. `r clim_sens`). 
This result, which is consistent with physiological expectations [**REFS**, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear. 
The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].

>>>>

Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021], and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @rollinson_climate_2021], which may often be best described by nonlinear functions [REFS, @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., @evans_fusing_2017; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously address all the questions posed here (Table `r hypotheses`).

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
Over the vast majority of Earth's forested regions, water is the primary climatic factor limiting tree growth, whereas low temperatures are the strongest limitation in some high latitude or high elevation sites [@babst_twentieth_2019]. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the main growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., **REFS**, @buntgen_2500_2011], more in-depth characterization of climate sensitivity across a representative sample of all trees in a forest stand is desirable for efforts to understand forest productivity and its climate responses [@babst_when_2018].
First, the additive or interactive effects of moisture and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020].
While drought metrics such as the Standardised Precipitation-Evapotranspiration Index [SPEI, @vicente-serrano_multiscalar_2010] integrate effects of precipitation and temperature, soil moisture and atmospheric vapor pressure deficit (VPD) affect plant growth differently [@novick_increasing_2016], and we hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`).
In addition, based on the predominance of nonlinear climate responses of tree growth over both smaller and larger spatiotemporal scales [e.g., @kumarathunge_acclimation_2019; @banburymorgan_global_2021], we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @trouillier_size_2019; @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].

@@

Here, we develop, and apply to ten forested sites spanning 52 degrees latitude, a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
This approach allows us to address the following broad questions concerning and associated hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) for a globally distributed tree-ring dataset:
(1) How does interannual climate variation shape annual tree growth?
(2) How does growth rate vary with DBH?
(3) How have size-corrected growth rates changed through time?

@@

Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability (PPT or PDF) -- at least up until the long-term mean -- 
and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
These findings are generally consistent with current understanding of global-scale patterns in climate sensitivity [@rozendaal_dendroecology_2011; @babst_twentieth_2019]: the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone). 
Within warmer regions, forests in humid climates can sometimes benefit from warm winter or spring temperatures [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig. `r clim_sens`).
However, the predominantly negative temperature responses (Fig. `r clim_sens`) imply that warmer temperatures are likely to reduce growth across the wide range of forest types and climates represented here.
This growth decrease at high temperatures is probably driven primarily by increased atmospheric dryness (PET or VPD) and ensuant exacerbation of observed water limitations [@novick_increasing_2016; @humphrey_soil_2021; @lopez_systemic_2021].
This effect occurs in addition to the effects of precipitation (Fig. `r clim_sens`), highlighting the fact that temperature and associated VPD increases limit growth even under conditions of high soil moisture [@novick_increasing_2016], and occurs over shorter time-frames (usually $\le$ 3 mo) than the effects of precipitation (usually $\ge$ 3 mo.; Table `r hypotheses`, Fig. `r clim_sens`).
This suggests that relatively short periods of anomalously high temperatures and atmospheric dryness, themselves caused in large part by soil dryness [@humphrey_soil_2021], add to effects of prolonged periods of reduced precipitation to shape forest drought responses.

Our analysis differed fundamentally from most conventional approaches in testing for non-linear responses of growth to climate, finding that the majority of climate responses were nonlinear (Table `r hypotheses`, Fig. `r clim_sens`). 
This result, which is consistent with physiological expectations [**REFS**, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear. 
The nonlinear form of most climate growth responses implies that as the climate changes such that high temperatures and strong precipitation anomalies become more common [@ipcc_climate_2014], non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].
